Model based reinforcement learning at FERMI FEL

### 2020-11-17 main.tex
txs auto checkin

\begin{abstract}
In this paper we discuss a model based reinforcement learning approach in comparison to a model free reinforcement learning approach applied at the FERMI FEL system. Both algorithms and approaches are new in this context and the main purpose of this paper is to show that reinforcement learning can be applied on an operational level in a feasible training time on real accelerator physics problems. In terms of sample-complexity the model-based approach is faster, while the finale performance of the model free method is superior - the so called asymptotic performance. The model-based algorithm is done in a Dyna-style using an uncertainty aware model and the model-free algorithm is based on tailored deep Q-learning using some tricks to increase the sample efficiency
\end{abstract}
%\keywords{Suggested keywords}%Use showkeys class option if keyword
                              %display desired
\maketitle
\begin{itemize}
	\item Noise benchmarks with pendulum
	\item Benchmark NAF2 with pendulum

\end{itemize}
%\tableofcontents
\section{Introduction and Motivation}
In particle accelerators one main goal is to provide stable and reproducible performance. In order to achieve this, a number of control problems have to be considered simultaneously. Especially if there is no way to model the physics, one might use optimisation techniques as e.g. derivative free optimisers (DFOs) or model-based optimisations as Gaussian processes. Another promising way is to apply reinforcement learning which has several advantages over optimisation methods:
\begin{itemize}
    \item Covers a larger class of problems.
    \item Solves the problem not always completely from the beginning as an DFO.
    \item Ability to use existing data.
    \item The underlying structure of the problem might be deduced.
\end{itemize}
RL optimises a sequence of decisions. On critical aspect using RL is the number of iteration to train a controller.
In this paper, we present the study carried out to solve the maximisation-problem of the radiation intensity generated by a seeded Free-Electron Laser (FEL) on the Free Electron laser Radiation for Multidisciplinary Investigations (FERMI) at Elettra Sincrotrone Trieste. Three algorithms were applied successfully showing different advantages, which will e discussed in the following.

@@

\section{Deep Reinforcement learning}
Assume states $s$ the set of all states $\bs$ and actions $\ba$ all actions $\actions$ and rewards $\reward$ and the set of all rewards $\mathcal{R}$ an initial state distribution $\initstate$.
Goal of reinforcement learning is to find a $\policy\mapsto a$, which is the solution of:
\begin{align}
\max J (\policy)  =  \mathbb E_{\traj \sim p_{\policy}}\left[\sum_{t=0}^{\horizon}\discount^t r(\bs_t,\ba_t)\right],
\end{align}
where $p_\policy = \initstate(\bs_0)\prod_{t_0}^\horizon\policy(\ba_t,\bs_t)\transitions(\bs_{t+1}|\bs_t,\ba_t)$ is the distribution of all trajectories $\traj := (\bs_0, \ba_0, \bs_1, a_1,\dots s_\horizon,a_\horizon)$ drawn by $\policy$ with a horizon $\horizon$.
In the modern field of RL one distinguishes if the policy $\policy(\ba)\approx \policy_\phi(\bs)$ or the state-action value function $Q(\bs,\ba)$ is approximated using a high capacity function approximator, as e.g. a deep neural network. In the first case one speaks about policy gradient methods in the latter about approximate dynamic programming, which we now discuss.
\subsection{Approximate dynamic programming}
The state-value function $Q(\bs,\ba)$ is expressed as $Q_\theta(\bs,\ba)$, where $\theta$ denotes the parameters of the function approximator. By satisfying the Bellmann-optimality operator the optimal $Q^*(\bs,\ba)$ can be found:
\begin{align}
    \min_\theta \left(\vec Q_\theta - \bellman^*\vec Q_\theta\right)^2.\label{eq:minimize_bellmann_optimality}
\end{align}
This operator has a unique fixed point but is non-linear, due to the $\max$ - operator:
\begin{align}
 \bellman^*\vec Q_\theta(\bs_t,\ba_t) := r(\bs_t,\ba_t)+\gamma\max_\ba\left( Q_\theta(\bs_{t+1},\ba) - Q_\theta(\bs_t,\ba_t)\right).
\end{align}
The form of this equation can cause overestimation and other complications, when using a function approximator. Several methods exist which try to mitigate the problems \cite{}.
One way to reduce the effect is to take a simple analytical form of the $Q$-function.
\subsection{NAF}
If a specific quadratic form of the $Q$ function is assumed:
\begin{align}
     Q_\theta(\bs,\ba) = -\frac{1}{2}(\ba-\mu_\theta(\bs))P_\theta(\ba-\mu_\theta(\bs))^T+V_\theta(s).
\end{align}
 The policy is given by:
\begin{align}
    \policy_\theta(\bs)=(\delta(\ba)-\arg\!\max_\ba Q_\theta(\bs,\ba)).
\end{align}
One modification, which is used in these tests is a twin network (weights denoted by $\tilde\theta$), where only one is used to obtain the policy, while the other is used for the update rule to avoid over-estimation.
The maximum is given analytically as $\max_a Q(s,a) = V(s)$, hence from \cref{eq:minimize_bellmann_optimality}:
\begin{align}
	\min_\theta\left( (\reward(\bs_t,\ba_t)+\gamma \min_{1,2} V_{\theta^i_\text{targ}}(\bs_{t+1}) - (1+\gamma) Q_\theta(\bs_t,\ba_t)\right)^2
\end{align}
$\theta_\text{targ}$ are the weights of a target network, which is softly updated. To stabilize the network training a small artificial noise is added to the actions in the update. This algorithm has an extremely good sample-efficiency for suitable problems as can be found in accelerators and is used as the baseline for the considered control problem.
%\subsection{New NAF network architecture}

@@

\appendix
\section{A non-linear standard control problem}
To provide some transparency of these studies we provide results on a famous classical standard control problem, the inverted pendulum. It is a non-linear low dimensional unsolved continuous control problem, which means there is no threshold for the reward to terminate an episode. The episode length is set to 200 steps.
\section{NAF2 details}\label{appendix:pernaf}

>>>>

\begin{abstract}
In this paper we discuss a model based reinforcement learning approach in comparison to a model free reinforcement learning approach applied at the FERMI FEL system. Both algorithms and approaches are new in this context and the main purpose of this paper is to show that reinforcement learning can be applied on an operational level in a feasible training time on real accelerator physics problems. In terms of sample-complexity the model-based approach is faster, while the finale performance of the model free method is superior - the so called asymptotic performance. The model-based algorithm is done in a Dyna-style using an uncertainty aware model and the model-free algorithm is based on tailored deep Q-learning using some tricks to increase the sample efficiency
\end{abstract}
%\keywords{Suggested keywords}%Use showkeys class option if keyword
                              %display desired
\maketitle
\begin{itemize}
	\item Noise benchmarks with pendulum
	\item Benchmark NAF2 with pendulum
	\item What is our improvement concerning the algorithm
	\item The first time model based on accelerator problem, 
\end{itemize}
%\tableofcontents
\section{Introduction and Motivation}
In particle accelerators one main goal is to provide stable and reproducible performance. In order to achieve this, a number of control problems have to be considered simultaneously. Especially if there is no way to model the physics, one might use optimisation techniques as e.g. derivative free optimisers (DFOs) or model-based optimisations as Gaussian processes. Another promising way is to apply reinforcement learning which has several advantages over optimisation methods:
\begin{itemize}
    \item Covers a larger class of problems.
    \item Solves the problem not always completely from the beginning as an DFO.
    \item Ability to use existing data.
    \item The underlying structure of the problem might be deduced.
\end{itemize}
RL optimises a sequence of decisions. On critical aspect using RL is the number of iteration to train a controller.
In this paper, we present the study carried out to solve the maximisation-problem of the radiation intensity generated by a seeded Free-Electron Laser (FEL) on the Free Electron laser Radiation for Multidisciplinary Investigations (FERMI) at Elettra Sincrotrone Trieste. Three algorithms were applied successfully showing different advantages, which will e discussed in the following.

@@

\section{Deep Reinforcement learning}
Assume states $s$ the set of all states $\bs$ and actions $\ba$ all actions $\actions$ and rewards $\reward$ and the set of all rewards $\mathcal{R}$ an initial state distribution $\initstate$.
Goal of reinforcement learning is to find a $\policy\mapsto a$, which is the solution of:
\begin{align}
\max J (\policy)  =  \mathbb E_{\traj \sim p_{\policy}}\left[\sum_{t=0}^{\horizon}\discount^t r(\bs_t,\ba_t)\right],
\end{align}
where $p_\policy = \initstate(\bs_0)\prod_{t_0}^\horizon\policy(\ba_t,\bs_t)\transitions(\bs_{t+1}|\bs_t,\ba_t)$ is the distribution of all trajectories $\traj := (\bs_0, \ba_0, \bs_1, a_1,\dots s_\horizon,a_\horizon)$ drawn by $\policy$ with a horizon $\horizon$.
In the modern field of RL one distinguishes if the policy $\policy(\ba)\approx \policy_\phi(\bs)$ or the state-action value function $Q(\bs,\ba)$ is approximated using a high capacity function approximator, as e.g. a deep neural network. In the first case one speaks about policy gradient methods in the latter about approximate dynamic programming, which we now discuss.
\subsection{Approximate dynamic programming}
The state-value function $Q(\bs,\ba)$ is expressed as $Q_\theta(\bs,\ba)$, where $\theta$ denotes the parameters of the function approximator. By satisfying the Bellmann-optimality operator the optimal $Q^*(\bs,\ba)$ can be found:
\begin{align}
    \min_\theta \left(\vec Q_\theta - \bellman^*\vec Q_\theta\right)^2.\label{eq:minimize_bellmann_optimality}
\end{align}
This operator has a unique fixed point but is non-linear, due to the $\max$ - operator:
\begin{align}
 \bellman^*\vec Q_\theta(\bs_t,\ba_t) := r(\bs_t,\ba_t)+\gamma\max_\ba\left( Q_\theta(\bs_{t+1},\ba) - Q_\theta(\bs_t,\ba_t)\right).
\end{align}
The form of this equation can cause overestimation and other complications, when using a function approximator. Several methods exist which try to mitigate the problems \cite{}.
One way to reduce the effect is to take a simple analytical form of the $Q$-function.
\subsection{NAF}
If a specific quadratic form of the $Q$ function is assumed:
\begin{align}
     Q_\theta(\bs,\ba) = -\frac{1}{2}(\ba-\mu_\theta(\bs))P_\theta(\ba-\mu_\theta(\bs))^T+V_\theta(s).
\end{align}
 The policy is given by:
\begin{align}
    \policy_\theta(\bs)=(\delta(\ba)-\arg\!\max_\ba Q_\theta(\bs,\ba)).
\end{align}
One modification, which is used in these tests is a twin network (weights denoted by $\tilde\theta$), where only one is used to obtain the policy, while the other is used for the update rule to avoid over-estimation.
The maximum is given analytically as $\max_\ba Q(\bs,\ba) = V(\bs)$, hence from \cref{eq:minimize_bellmann_optimality}:
\begin{align}
	\min_\theta\left( (\reward(\bs_t,\ba_t)+\gamma \min_{1,2} V_{\theta^i_\text{targ}}(\bs_{t+1}) - (1+\gamma) Q_\theta(\bs_t,\ba_t)\right)^2
\end{align}
$\theta_\text{targ}$ are the weights of a target network, which is softly updated. To stabilize the network training a small artificial noise is added to the actions in the update. This algorithm has an extremely good sample-efficiency for suitable problems as can be found in accelerators and is used as the baseline for the considered control problem.
%\subsection{New NAF network architecture}

@@

\appendix
\section{A non-linear standard control problem}
To provide some transparency of these studies we provide results on a famous classical standard control problem, the inverted pendulum. It is a non-linear low dimensional unsolved continuous control problem, which means there is no threshold for the reward to terminate an episode. The episode length is set to 200 steps.
\subsection{The impact of noise}
\section{NAF2 details}\label{appendix:naf2}
\section{AE training details}
\section{Theoretical aspects on the AE-DYNA on FERMI FEL}
